### 2025-03-14 paper/quarto_ecml/paper.tex
Fix indentation in hyperparameter results and discussion

We test the impact of three types of hyperparameters; our complete
results are in the supplementary appendix.\\
\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
Third, increasing \(T\), the maximum number of steps, generally yields
better outcomes because more CEs can mature in each training epoch.
Fourth, the impact of \(\tau\), the required decision threshold is more
difficult to predict. On ``harder'' datasets it may be difficult to
satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
increasing this threshold does not seem to correlate with better
outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
leads to optimal results because it is associated with high proportions
of mature counterfactuals.\\
\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

As our results indicate, counterfactual training produces models that
are more explainable. However, our approach is not without
limitations.\\
\textbf{\emph{CT increases the training time of models.}} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\textbf{\emph{Immutable features may have proxies.}} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to fulfill its intended goal.\\
\textbf{\emph{Interventions on features may impact fairness.}} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features, which may have implication for
the fair and equitable treatment of decision subjects. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}. Still,
these phenomena are not specific to CT.

We also highlight several interesting directions for future research.\\
\textbf{\emph{Extending CT beyond classification settings.}} Our
formulation relies on the distinction between non-target class(es)
\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
discrete. Thus, it does not apply to ML tasks where the change in
outcome cannot be readily quantified. Focus on classification models is
a common restriction in research on CEs and AR. Other settings have
attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of CEs.\\
\textbf{\emph{Addressing the training instabilities.}} JEMs are
susceptible to instabilities during training \cite{grathwohl2020your}
and even though we depart from the SGLD-based sampling, we still
encounter major variability in the outcomes. CT is exposed to two
potential sources of instabilities: (1) the energy-based contrastive
divergence term in Equation~\ref{eq-div}, and (2) the underlying
counterfactual explainers. Still, we find that training instabilities
can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each training epoch, and including only mature counterfactuals
for contrastive divergence.\\
\textbf{\emph{Improving hyperparameter selection procedures.}} Our
method benefits from the tuning of certain key hyperparameters (see
Section~\ref{sec-hyperparameters}). In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, CT is iterative which makes a variety of
methods applicable, including Bayesian \cite{snoek2012practical} or
gradient-based \cite{franceschi2017forward} optimization.

>>>>

We test the impact of three types of hyperparameters; our complete
results are in the supplementary appendix.\\
\indent \emph{Hyperparameters of the CE generators.} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
Third, increasing \(T\), the maximum number of steps, generally yields
better outcomes because more CEs can mature in each training epoch.
Fourth, the impact of \(\tau\), the required decision threshold is more
difficult to predict. On ``harder'' datasets it may be difficult to
satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
increasing this threshold does not seem to correlate with better
outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
leads to optimal results because it is associated with high proportions
of mature counterfactuals.\\
\indent \emph{Hyperparameters for penalties.} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes
(especially) when combined with high values of \(\lambda_{\text{div}}\)
or \(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
\indent \emph{Other hyperparameters.} We observe that the effectiveness
and stability of CT is positively associated with the number of
counterfactuals generated during each training epoch. We also confirm
that a higher number of training epochs is beneficial. Interestingly, we
observed desired improvements in explainability when CT was combined
with conventional training and applied only for the final 50\% of epochs
of the complete training process. Put differently, CT may be a way to
improve the explainability of models in a fine-tuning manner.

@@

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature \cite{venkatasubramanian2020philosophical} impacting
the capacity of CT to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. As protecting a set
of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}. Still,
these phenomena are not specific to CT.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily quantified. Focus on
classification models is a common restriction in research on CEs and AR.
Other settings have attracted some interest (e.g., regression in
\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
little consensus how to robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \cite{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian (e.g., \cite{snoek2012practical}) or gradient-based (e.g.,
\cite{franceschi2017forward}) optimization applicable.

### 2025-03-14 paper/sections/discussion.qmd
Fix indentation in hyperparameter results and discussion

As our results indicate, counterfactual training produces models that are more explainable. However, our approach is not without limitations.  
***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

We also highlight several interesting directions for future research.  
***Extending CT beyond classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
***Addressing the training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.  
***Improving hyperparameter selection procedures.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

>>>>

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Still, these phenomena are not specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian (e.g., [@snoek2012practical]) or gradient-based (e.g., [@franceschi2017forward]) optimization applicable.

### 2025-03-14 paper/sections/experiments.qmd
Fix indentation in hyperparameter results and discussion

### Impact of hyperparameter settings. {#sec-hyperparameters}
We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.  
***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.  
***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

>>>>

### Impact of hyperparameter settings. {#sec-hyperparameters}
We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
`\indent`{=latex} *Hyperparameters of the CE generators.* First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time. Second, we note that the choice of a CE generator has a major impact on the results. For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful]. Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch. Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict. On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes. In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.  
`\indent`{=latex} *Hyperparameters for penalties.* We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes (especially) when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.   
`\indent`{=latex} *Other hyperparameters.* We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.
